---
title: "3_Figures_20260202"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(terra)
library(tidyverse)
library(ggplot2)  
library(scales)
library(sf)
library(tigris)
library(ggspatial)
library(ggpubr)
library(cowplot)
library(patchwork)
library(ggh4x)
library(grid)
library(gridExtra)
library(png)
library(ggrepel)
library(paletteer)
library(biscale)
library(leaflet)
library(viridisLite)

## upload necessary data
ca_shp <- counties(state = "CA", cb = TRUE, class = "sf", year = 2022)
```

#Main Text

##Figure 1. Conceptual diagram

part a. People
```{r}
## upload population density
gpw <- rast("data/raw/gpw-v4-population-density-rev11_2020_30_sec_asc/gpw_v4_population_density_rev11_2020_30_sec_1.asc")

# transform to ca crs
ca_shp_gpw <- ca_shp %>% 
  st_transform(crs(gpw)) %>% 
  vect() %>% 
  {mask(crop(gpw, .), .)} # new trick, make intermediate objects instead of new mask, crops, etc
#ca_shp_gpw[is.na(ca_shp_gpw)] <- 0 
gpw_df <- as.data.frame(ca_shp_gpw, xy = TRUE)
colnames(gpw_df)[3] <- "pop_density"

# to help with log10 transformation
gpw_df$pop_density <- ifelse(gpw_df$pop_density == 0, 0.001, gpw_df$pop_density)

map_popdensity <- ggplot() +
  geom_raster(data = gpw_df, aes(x = x, y = y, fill = pop_density)) +
  geom_sf(data = ca_shp, fill = NA, color = "black", size = 0.6) +
  scale_fill_viridis_c(trans = "log10", na.value = "grey95", option = "mako", direction = -1,
                       labels = label_number(accuracy = 0.001, scale_cut = cut_short_scale())) +
  coord_sf() +
  theme_void() +
  labs(title = "PEOPLE", subtitle = "Population Density", fill = "People/km²") + 
  theme(plot.title = element_text(face = "bold", size = 15), 
        plot.subtitle = element_text(size = 14),
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.key.width = unit(1, "cm"), 
        legend.key.height = unit(.25, "cm"),
        legend.title = element_text(size = 14), #vjust = .7, 
        legend.text = element_text(size = 12),
        legend.box.spacing = unit(0,"pt")) +
  guides(fill = guide_colorbar(title.position = "top", title.hjust = 0))
```

part b. Environment
```{r}
tick_inf <- rast("data/raw/ticks/ipac_year_infection_predictions.tif") %>% 
  project("EPSG:4269") 


ca_shp_tick_inf <- ca_shp %>% 
  st_transform(crs(tick_inf)) %>% 
  vect() %>% 
  {mask(crop(tick_inf, .), .)}

tick_inf_df_agg <- aggregate(tick_inf, fact = 5, fun = mean)
tick_inf_df_agg_df <- as.data.frame(tick_inf_df_agg, xy = TRUE, na.rm = TRUE)
colnames(tick_inf_df_agg_df)[3] <- "tick_inf"
tick_inf_df_agg <- aggregate(tick_inf, fact = 5, fun = mean)
#write.csv(tick_inf_df, file = "tick_inf_df.csv")

map_tickinf <- ggplot() +
  geom_raster(data = tick_inf_df_agg_df, aes(x = x, y = y, fill = tick_inf)) +
  geom_sf(data = ca_shp, fill = NA, color = "black", size = 0.6) +
  scale_fill_viridis_c(na.value = "grey95", option = "mako", direction = -1) +
  coord_sf() +
  theme_void() +
  labs(title = "ENVIRONMENT", subtitle = "Infected Ticks", fill = "Probability") + 
  theme(plot.title = element_text(face = "bold", size = 15), 
        plot.subtitle = element_text(size = 14),
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.key.width = unit(1, "cm"), 
        legend.key.height = unit(.25, "cm"),
        legend.title = element_text(size = 14), #vjust = .7, 
        legend.text = element_text(size = 12),
        legend.box.spacing = unit(0,"pt")) +
  guides(fill = guide_colorbar(title.position = "top", title.hjust = 0))


```
part c. Exposure
```{r}
## fix ca_shp to filter larger shp files
ca_bbox <- st_bbox(ca_shp) 
ca_bbox_sf <- st_as_sfc(ca_bbox)

## read in park boundaries
casp <- st_read("data/raw/parks/casp_individualpark_boundaries/ParkBoundaries.shp") %>% 
  st_transform(st_crs(ca_shp)) 

us_nps <- st_read("data/raw/parks/nps_boundary/nps_boundary.shp",
                  wkt_filter = st_as_text(ca_bbox_sf)) %>% # filter to ca before or else it's really slow
  st_transform(st_crs(ca_shp)) %>% 
  st_make_valid() %>% 
  st_intersection(ca_shp)

us_nfs <- st_read("data/raw/parks/nsf_boundary/S_USA.BdyAdm_LSRS_AdministrativeForest.shp",
                  wkt_filter = st_as_text(ca_bbox_sf)) %>% # filter to ca before or else it's really slow
  st_transform(st_crs(ca_shp)) %>% 
  st_make_valid() %>% 
  st_intersection(ca_shp)   # filter 
 


## plot it

map_recland <- ggplot() +
  geom_sf(data = ca_shp, fill = "grey90", color = "black", size = 0.5) +
  geom_sf(data = us_nfs, fill = "#008080", aes(color ="NFS"), alpha = 0.9, size = 0.5) +
  geom_sf(data = us_nps, fill = "goldenrod", aes(color = "NPS"), alpha = 0.9, size = 0.5) +
  geom_sf(data = casp, fill = "darkred", aes(color ="CASP"), alpha = 0.9, size = 0.5) +
  
   annotation_north_arrow(location = "tr", which_north = "true",
                         height = unit(1.2, "cm"),
                         width = unit(1.2, "cm"),
                         pad_x = unit(1.6, "cm")) + #, pad_y = unit(1, "cm") # 2.5 also worked
  
  annotation_scale(location = "tr", width_hint = 0.3, text_cex = 0.9,
                   pad_x = unit(1.6, "cm"), pad_y = unit(2, "cm")) +
 
  scale_color_manual(values = c("CASP" = "darkred",
                               "NPS" = "goldenrod",
                               "NFS" = "#008080"), name = "Management") +
  theme_void() +
  labs(title = "EXPOSURE", subtitle = "Recreation Land") + 
  theme(plot.title = element_text(face = "bold", size = 15), 
        plot.subtitle = element_text(size = 14),
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.box = "vertical",
        legend.title.position = "top",
        legend.key.width = unit(1, "cm"), 
        legend.key.height = unit(.25, "cm"),
        legend.title = element_text(size = 14), #vjust = .7, 
        legend.text = element_text(size = 12),
        legend.box.spacing = unit(0,"pt")) +
  guides(color = guide_legend(title.hjust = 0, direction = "horizontal")) 

```


Combine them
```{r}
map_combo <- ggarrange(map_popdensity, map_tickinf, map_recland,
          nrow = 1, ncol = 3, 
          widths = c(1,1,1), align = "hv")



ggsave(map_combo, file = "figures/updated/fig_1_v4.png", width = 10, height = 7, dpi = 300, bg = "white") # 10 8
```

##Figure 2. Where

part a. infected ticks
```{r}

## set up data
casp_tick_allextractions <- read.csv(file = "data/processed/casp_tick_allextractions.csv")  %>% 
  mutate(UNITNAME = case_when(UNITNAME == "Mount Diablo SP - Morgan Territory Regional Preserve" ~ "Mount Diablo SP ",
                              TRUE ~ UNITNAME))


casp_tick_allextractions_shp <- casp_shp_clean  %>% 
  mutate(UNITNAME = case_when(UNITNAME == "Mount Diablo SP - Morgan Territory Regional Preserve" ~ "Mount Diablo SP ",
                              TRUE ~ UNITNAME)) %>% 
  left_join(casp_tick_allextractions, by = "UNITNAME") %>% 
  st_transform(st_crs(ca_shp)) %>% 
  st_make_valid() %>% 
  st_centroid() %>% 
  mutate(label = UNITNAME,
         x = st_coordinates(.)[, "X"],
         y = st_coordinates(.)[, "Y"])   

casp_tick_top10 <- casp_tick_allextractions_shp  %>% 
  filter(UNITNAME != "Mount Diablo SP - Diablo Foothills Regional Park") %>%  #UNITNAME != "Mount Diablo SP - Morgan Territory Regional Preserve" & 
  group_by(UNITNAME, x, y, label) %>% 
  summarise(tick_year_inf_prob = max(tick_year_inf_prob), .groups = "drop") %>% 
  arrange(desc(tick_year_inf_prob))  %>% 
  slice_head(n = 10)

map_casptickinf <- 
  ggplot() +
  geom_sf(data = ca_shp, fill = "grey40", color = "black") +
  geom_sf(data = casp_tick_allextractions_shp, aes(color = tick_year_inf_prob), size = 2.5, alpha = .9) +
  geom_text_repel(data = casp_tick_top10, aes(x = x, y = y, label = label), size = 4.5, 
                  nudge_x = -7.7, direction = "y", hjust = 1, force = 3.3,box.padding = 0.6,  
                  segment.color = "gray50", segment.size = 0.3, max.overlaps = 20) +
  scale_color_paletteer_c("grDevices::YlOrRd", direction = -1, na.value = "transparent")  +
  theme_void() +
  labs(title = "Infected Ticks", color = "Probability") + #subtitle = "Top 10 Parks", 
    theme(plot.title = element_text(face = "bold", size = 15), 
        #plot.subtitle = element_text(size = 14),
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.box = "vertical",
        legend.title.position = "top",
        legend.key.width = unit(1, "cm"), 
        legend.key.height = unit(.25, "cm"),
        legend.title = element_text(size = 14), #vjust = .7, 
        legend.text = element_text(size = 12),
        legend.box.spacing = unit(0,"pt")) +
  guides(color = guide_colorbar(title.hjust = 0, direction = "horizontal")) 
  



```

part b. Human Activity
```{r}
casp_year_total <- read.csv(file = "data/processed/casp_year_totalvisitors.csv")


casp_visit_allextractions_shp <- casp_shp_clean  %>% 
  left_join(casp_year_total, by = "UNITNAME") %>% 
  st_transform(st_crs(ca_shp)) %>% 
  st_make_valid() %>% 
  st_centroid() %>% 
  mutate(label = UNITNAME,
         x = st_coordinates(.)[, "X"],
         y = st_coordinates(.)[, "Y"])   %>% 
  mutate(casp_year_meantotal = ifelse(is.na(casp_year_meantotal), 1, casp_year_meantotal),
         casp_year_meantotal = ifelse(is.na(casp_year_meantotal) | casp_year_meantotal == 0, 1, casp_year_meantotal))


casp_visit_top10 <- casp_visit_allextractions_shp   %>% 
  group_by(UNITNAME, x, y, label) %>% 
  summarise(casp_year_meantotal = max(casp_year_meantotal), .groups = "drop") %>% 
  arrange(desc(casp_year_meantotal))  %>% 
  slice_head(n = 13)   %>%  # need to get rid of duplicates
  group_by(label) %>% 
  slice(1) %>% 
  filter(label != "South Carlsbad SB")
  

map_caspvisits <- 
ggplot() +
  geom_sf(data = ca_shp, fill = "grey40", color = "black") +
  geom_sf(data = casp_visit_allextractions_shp, aes(color = casp_year_meantotal), size = 2.5, alpha = .9) +
  geom_text_repel(data = casp_visit_top10, aes(x = x, y = y, label = label), size = 4.5, 
                  nudge_x = -7.5, direction = "y", hjust = 1, force = 4,box.padding = 0.6,  
                  segment.color = "gray50", segment.size = 0.3, max.overlaps = 20) +
    scale_color_paletteer_c("grDevices::YlOrRd", direction = -1, na.value = "transparent", trans = "log10", 
                        labels = label_number(accuracy = 1, scale_cut = cut_short_scale()))  +
  theme_void() +
  labs(title = "Annual Visitors", color = "log(Total)") + #subtitle = "Top 10 Parks", 
    theme(plot.title = element_text(face = "bold", size = 15), 
        #plot.subtitle = element_text(size = 14),
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.box = "vertical",
        legend.title.position = "top",
        legend.key.width = unit(1, "cm"), 
        legend.key.height = unit(.25, "cm"),
        legend.title = element_text(size = 14), #vjust = .7, 
        legend.text = element_text(size = 12),
        legend.box.spacing = unit(0,"pt")) +
  guides(color = guide_colorbar(title.hjust = 0, direction = "horizontal")) 



```

combine them
```{r}
map_combo_fig2 <- ggarrange(map_casptickinf, map_caspvisits,
          nrow = 1, ncol = 2,
          widths = c(2,2), align = "hv")


ggsave(map_combo_fig2, file = "figures/updated/fig_2_v3.png", width = 10, height = 7, dpi = 300, bg = "white") # 10 8

```

##Figure 3. When

part a. Biscale figure of parks
```{r}
## merge distance with sp data
min_dist_km <- read.csv("data/processed/casp_dist_coast_km.csv") %>% 
  dplyr::select(dist_coast_km) %>% #UNITNAME,
  mutate(dist_coast_km = as.numeric(dist_coast_km)) 

# probs should do with _select parks first but will do it this way for now
casp_shp$dist_coast_km <- min_dist_km$dist_coast_km 

casp_shp_dist <- casp_shp %>% # 462 original parks
  filter(grepl(" SP", UNITNAME) |
           grepl(" SRA", UNITNAME) |
           grepl(" SB", UNITNAME) |
           grepl(" SNR", UNITNAME) |
           grepl(" SW", UNITNAME))   %>% 
  st_transform(st_crs(ca_shp)) %>%
  st_point_on_surface() %>% 
  st_transform(crs = st_crs(ca_shp)) %>% 
  {cbind(., st_coordinates(.))} %>% 
  dplyr::select(UNITNAME, dist_coast_km, X, Y, geometry)

## create biscale
# quantile
# bc ensures roughly equal numbers of parks per class, emphasizes relative position along gradients
casp_shp_biscale <- bi_class(casp_shp_dist, x = dist_coast_km, y = Y, style = "equal", dim = 3) # "equal"/quantile


biscale_colors <- bi_pal(pal = "DkViolet", dim = 3, preview = FALSE)

## plot
# base map
bivar_map <- ggplot() +
  geom_sf(data = ca_shp, fill = NA, color = "black", size = 0.6) +
  geom_sf(data = casp_shp_biscale, aes(color = bi_class), size = 2, show.legend = FALSE, alpha = .9) +
  bi_scale_color(pal = "DkViolet", dim = 3) +
  bi_theme() +
  theme_void()


# legend
bivar_legend <- bi_legend(pal = "DkViolet", dim = 3,
                          xlab = "Dist. from Coast", ylab = "Rel. Latitude",
                          pad_width = 0.05, size = 16)

# combin
map_casp_bivar <-ggdraw() +
  draw_plot(bivar_map, x = -0.15, y = -0.051, width = 1.3, height = 1.1) +
  draw_plot(bivar_legend, x = 0.47, y = 0.65, width = 0.39, height = 0.39)


ggsave(map_casp_bivar, file = "figures/updated/map_casp_bivar_v4.png", dpi = 600, width = 4.2, height = 4)

```

another version of legend map (not sure if this is better)
```{r}
# project CA counties
ca_shp_proj <- st_transform(ca_shp, 3310)

# make grid (10 km)
ca_grid <- st_make_grid(
  ca_shp_proj,
  cellsize = 10000,
  square = TRUE
) %>%
  st_sf() %>%
  st_intersection(ca_shp_proj)

# coastline
coast <- ne_download(
  scale = 50,
  type = "coastline",
  category = "physical",
  returnclass = "sf"
) %>%
  st_transform(3310)

# distance to coast (km)
ca_grid$dist_coast_km <- as.numeric(
  st_distance(ca_grid, coast) |> apply(1, min)
) / 1000

# latitude proxy from centroids (DO NOT overwrite geometry)
coords <- st_coordinates(st_centroid(ca_grid))
ca_grid$Y <- coords[,2]

# standardize gradients
ca_grid <- ca_grid %>%
  mutate(
    coast_z = scale(dist_coast_km)[,1],
    lat_z   = scale(Y)[,1]
  )

# biscale grid polygons
ca_grid_bi <- bi_class(
  ca_grid,
  x = coast_z,
  y = lat_z,
  style = "quantile",
  dim = 3
)

# park points
casp_pts <- casp_shp %>%
  st_transform(3310) %>%
  st_point_on_surface()

# named bivariate palette (THIS IS THE FIX)
ca_grid_bi$bi_class <- factor(
  ca_grid_bi$bi_class,
  levels = as.vector(outer(1:3, 1:3, paste, sep = "-")))

bi_cols <- bi_pal(pal = "DkViolet", dim = 3)
bi_cols <- as.vector(unname(bi_cols[, ]))
names(bi_cols) <- levels(ca_grid_bi$bi_class)

# plot
ggplot() +
  geom_sf(data = ca_grid_bi, aes(fill = bi_class), color = NA) +
  geom_sf(data = casp_pts, color = "black", size = 0.6, alpha = 0.8) +
  #scale_fill_manual(values = bi_class, na.value = "transparent") +
  coord_sf() +
  theme_void() +
  bi_scale_fill(pal = "DkViolet", dim = 3) 



 #6. Create bivariate class
ca_grid_bi <- bi_class(
  ca_grid,
  x = coast_z,
  y = lat_z,
  style = "quantile",
  dim = 3
)

# 7. Park points (projected + safe centroid)
casp_pts <- casp_shp %>%
  st_transform(3310) %>%
  st_point_on_surface()

# 8. Plot map with bivariate grid, parks, and legend
bivar_map2 <- ggplot() +
  geom_sf(data = ca_grid_bi, aes(fill = bi_class), color = NA) +
  geom_sf(data = ca_shp_proj, fill = NA, color = "grey50") +
  geom_sf(data = casp_pts, color = "black", size = 0.8, alpha = 0.9) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  coord_sf() +
  theme_void() +
  guides(fill = FALSE)


# base map
bivar_map <- ggplot() +
  geom_sf(data = ca_shp, fill = NA, color = "black", size = 0.6) +
  geom_sf(data = casp_shp_biscale, aes(color = bi_class), size = 2, show.legend = FALSE, alpha = .9) +
  bi_scale_color(pal = "DkViolet", dim = 3) +
  bi_theme() +
  theme_void()


# legend
bivar_legend <- bi_legend(pal = "DkViolet", dim = 3,
                          xlab = "Dist. from Coast", ylab = "Rel. Latitude",
                          pad_width = 0.05, size = 16)

# combin
map_casp_bivar <-ggdraw() +
  draw_plot(bivar_map2, x = -0.15, y = -0.051, width = 1.3, height = 1.1) +
  draw_plot(bivar_legend, x = 0.47, y = 0.65, width = 0.39, height = 0.39)
```

part b. Tick seasonality
```{r}

casp_tick_monthly_suit_long <- read.csv("data/processed/casp_tick_monthly_suit_long.csv")

casp_tick_monthly_suit_distinct <- casp_tick_monthly_suit_long  %>%
  mutate(month = factor(month, levels = month.abb),
         month_num = as.numeric(month)) %>% 
  dplyr::select(UNITNAME, month_num,tick_monthly_suitability) %>% 
  rename(tick_monthly_presence = tick_monthly_suitability,
         month = month_num) %>% 
  distinct(UNITNAME, month, .keep_all = TRUE) 


casp_tick_monthly_suit_distinct_sf <- casp_tick_monthly_suit_distinct %>% 
  left_join(casp_shp_biscale %>% st_drop_geometry() %>% dplyr::select(UNITNAME, bi_class), by = "UNITNAME") 

fig_3_b <- casp_tick_monthly_suit_distinct_sf %>% 
ggplot(aes(x = as.factor(month), y = tick_monthly_presence)) +
  geom_boxplot(outlier.shape = NA, fill = "grey95") +
  geom_jitter(aes(color =  bi_class), alpha = .7, width = 0.1 ) +
  scale_color_manual(values = biscale_colors) +
  theme_classic() +
  scale_x_discrete(labels = month.abb) +
  guides(color = FALSE) +
  labs(y = "Tick Habitat Suitability") + # change to suitability
    theme(#plot.title = element_text(face = "bold", size = 14), #hjust = 0.5,
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        axis.title.y = element_text(face = "bold", size = 18),
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        legend.position = "bottom",
        legend.direction = "horizontal", 
        legend.key.width = unit(1, "cm"), 
        legend.title = element_text(vjust = .7))

```


part c. Human activity
```{r}
bi_class_levels <- c("1-3", "2-3", "3-3",
                     "1-2", "2-2", "3-2",
                     "1-1", "2-1", "3-1")

facet_label <- c("1-3" = "Coastal-North", 
                 "2-3" = "Mid-North", 
                 "3-3" = "Inland-North",
                 "1-2" = "Coastal-Mid", 
                 "2-2" = "Mid-Mid", 
                 "3-2" = "Inland-Mid",
                 "1-1" = "Coastal-South", 
                 "2-1" = "Mid-South", 
                 "3-1" ="Inland-South")

biscale_colors <- c(
  "1-3" = "#4885c1",  
  "2-3" = "#435786",  
  "3-3" = "#3f2949" , 
  "1-2" = "#89a1c8",  
  "2-2" = "#806a8a",  
  "3-2" = "#77324c",  
  "1-1" = "#cabed0",  
  "2-1" = "#bc7c8f",  
  "3-1" = "#ae3a4e" )[bi_class_levels] 


monthly_props <- read.csv("data/processed/activity_monthly_sourceprops_wide.csv") 

fig_3_c <- monthly_props %>%
  left_join(casp_shp_clean %>% st_drop_geometry(), by = "UNITNAME") %>%
  left_join(casp_shp_biscale %>% st_drop_geometry(), by = "UNITNAME") %>%
  mutate(bi_class = factor(bi_class, levels = c(
      "1-3", "2-3", "3-3",   # Top row = North
      "1-2", "2-2", "3-2",   # Middle row = Mid
      "1-1", "2-1", "3-1")))  %>% 
  mutate(source = recode(source,
                         "flickr_monthly_uniquetotals" = "Flickr",
                         "gbif_monthly_uniquetotals"   = "GBIF",
                         "meta_monthly_uniquetotals"   = "Meta"),
         source = factor(source, levels = c("Meta","Flickr", "GBIF"))) %>%
  ggplot(aes(x = month, fill = source)) +
  geom_density(aes(weight = monthly_prop), alpha = 0.5, color = NA, adjust = 1.2) +
  scale_x_continuous(breaks = 1:12, labels = substr(month.abb, 1, 1), name = "Month") +
  scale_fill_manual(values = c("Flickr" = "#D55E00", "GBIF" ="#E69F00", Meta = "#56B4E9"))+
  theme_classic() +
  labs(y = "Human Activity (Prop.)", fill = "Data Source") +
  theme(
    plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
    axis.title.y = element_text(face = "bold", size = 18),
    axis.title.x = element_blank(),
    axis.text.y = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    legend.position = "right",
    legend.text = element_text(size = 16),
    legend.direction = "vertical",
    legend.key.width = unit(1, "cm"),
    legend.title = element_text(vjust = .7, face = "bold", size = 18))  +
    facet_wrap2(~bi_class, nrow = 3, labeller = labeller(bi_class = facet_label),
               strip = strip_themed(background_x = elem_list_rect(fill = biscale_colors), 
                                    text_x = elem_list_text(face = "bold", color = "white", size = 14))) 


ggsave(fig_3_c, file = "figures/updated/fig_3bottom_v2.png", width = 16, height = 6, bg = "white", dpi = 300)
```

#Supplemental 

##Supplemental Figure 1. Maps of park adminstrative units
```{r}
## create centroids
casp_shp_select_centroids <- casp_shp_select %>% 
  st_centroid() %>% 
  st_transform(crs = st_crs(ca_shp)) %>% 
  {cbind(., st_coordinates(.))} %>% 
  filter(Shape_Area > 4800954) # median value

## version 2
casp_shp_select_centroids2 <- casp_shp_select %>% 
  st_centroid() %>% 
  st_transform(crs = st_crs(ca_shp)) %>% 
  {cbind(., st_coordinates(.))} %>% 
  filter(Shape_Area > quantile(Shape_Area, 0.90)) %>% # .75 top 25% by area
  filter(UNITNAME !="Santa Rosa Mountains SW")
  




map_casp_parks2 <-ggplot() +
  geom_sf(data = ca_shp, fill = "gray90", color = "gray70", size = 0.1) +
  geom_sf(data = casp_shp_select, fill =  "forestgreen", color = "forestgreen") +
  geom_text_repel(
    data = casp_shp_select_centroids2,
    aes(x = X, y = Y, label = UNITNAME),
    size = 3.5,
    box.padding = 0.5,
    point.padding = 0.3,
    force = 1.2,
    direction = "both",
    segment.color = "gray50",
    segment.alpha = 0.5,
    max.overlaps = Inf #Inf
  ) +
  theme_void() +
  #labs(title = "Select CA State Parks") +
  theme(plot.title = element_text(face = "bold", size = 16))

ggsave(map_casp_parks2 , file = "figures/fig_s_parkboundary_v4.png",  bg = "white")#width = 12, height = 15,
```


##Supplemental Figure 2. Park Admin boundaries
```{r}
## Field Division (n = 4)
map_casp_fielddivision <-  ggplot() +
    geom_sf(data = ca_shp, fill = NA, color = "gray50", size = 0.1) +
    geom_sf(data = casp_districtdivision_shp, aes(fill = field_divi), color = "black", alpha = 0.8) +
    theme_void() +
  labs(fill = "Field Division") +
  theme(legend.title = element_text(face = "bold")) +
  scale_fill_manual(values = c("#8CC8BCFF", "#EDB144FF","#C8570DFF",  "#7DA7EAFF"))

## District (n = 21)
base_palette <- paletteer_d("ggthemes::manyeys", n = 19)$colors
extended_palette <- c(base_palette,"#F4C2C2","#F0E5D8")#,"#F9D5B1")# "#E3CFC4" , "beige") #"#7F3C8D"

map_casp_district <-  
ggplot() +
    geom_sf(data = ca_shp, fill = NA, color = "gray50", size = 0.1) +
    geom_sf(data = casp_districtdivision_shp, aes(fill = district), color = "black", alpha = 0.8) +
    theme_void() +
  labs(fill = "District") +
  theme(legend.title = element_text(face = "bold")) +
  scale_fill_manual(values = extended_palette ) #+guides(fill = guide_legend(ncol = 1))
#legend_only <- get_legend(map_casp_district)
#png("map_casp_district_legend.png", width = 690, height = 1600, res = 300) # adjust size as needed
#grid::grid.draw(legend_only)
#dev.off()


## combine park admin
fig_s_parkadmin_boundary <- ggarrange(map_casp_fielddivision + labs(tag = "a"),map_casp_district+ labs(tag = "b"), 
                                      
                                      nrow = 2, ncol = 1, align = "hv", 
                                      widths = c(1,2), heights = c(.9,1) )
#ggsave(fig_s_parkadmin_boundary, file = "figures/fig_s_parkadmin_boundary.png", width = 6, height = 5.7, bg = "white")


ggsave(map_casp_fielddivision, file = "figures/fig_s_fielddivision.png", width = 6, height = 5.7, bg = "white")
ggsave(map_casp_district, file = "figures/fig_s_district.png", width = 6, height = 5.7,bg = "white")
```

##Supplemental Figure 3. Meta Activity Space Maps
```{r}

ca_shp <- counties(state = "CA", cb = TRUE, class = "sf", year = "2022")

all_tiles_sf_agg <- all_tiles_sf %>% 
  group_by(geom) %>% #xtile, ytile
  summarise(activity = mean(activity , na.rm = TRUE), .groups = "drop")


all_tiles_sf_agg <- st_transform(all_tiles_sf_agg, st_crs(casp_shp_clean))

all_tiles_ca <- st_intersection(
  all_tiles_sf_agg,
  st_make_valid(casp_shp_clean))


all_tiles_ca <- st_transform(all_tiles_ca, 4326)
casp_shp_clean      <- st_transform(casp_shp_clean, 4326)

pal <- colorNumeric(
  palette = viridis(256, option = "mako", direction = -1),
  domain  = log(all_tiles_ca$activity+1),
  na.color = "#f2f2f2")

leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles("CartoDB.Positron") %>%
  
  addPolygons(
    data = all_tiles_ca,
    fillColor   = ~pal(log(activity+1)),
    fillOpacity = 0.9,
    color       = NA,
    weight      = 0
  ) %>%
  
  addPolygons(
    data = casp_shp_clean,
    fill = FALSE,
    color = "black",
    weight = 1
  ) %>%
  
  addLegend(
    pal = pal,
    values = log(all_tiles_ca$activity+1),
    title = "Activity",
    position = "bottomright"
  )
```

##Supplemental Figure 4. Annual spatial covariates
part A. LD incidence
```{r}

ldi2 <- st_read("data/raw/LD_incidence/CA_Zips_Lyme_NAD83.shp") %>% 
  st_make_valid()

ldi2$LDI_Cat <- factor(ldi2$LDI_Cat, levels = sort(unique(ldi2$LDI_Cat))) 
  
ldi2 <- ldi2 %>% 
  mutate(LDI_Cat_Intervals = factor(LDI_Cat,
                                       levels = c(0, 1, 5, 10, 20, 50, 100),
                                       labels = c("0", "0.01 - 1.0", "1.01 -5.0", "5.1 -10.0",
                                                  "10.01 - 20.0", "20.01 - 50.0", ">50.0")))
map_ldi_zip <- 
  ggplot() +
  geom_sf(data = ldi2, aes(fill = LDI_Cat_Intervals), color = NA, size = 0.1) +
  scale_fill_viridis_d(option = "H", direction = 1, na.value = "grey90", name = "Lyme \nDisease\nIncidence") + 
  #labs(tag = "A") +
  theme_void() +
    theme(plot.title = element_text(face = "bold", size = 14), #hjust = 0.5,
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "right",#"bottom",# 
        legend.direction = "vertical", #"horizontal",# 
        #legend.key.width = unit(.25, "cm"), 
        #legend.key.height = unit(.25, "cm"),
        legend.title = element_text(vjust = .7,face = "bold")) +
  guides(fill = guide_legend(reverse = TRUE))

#ggsave(map_ldi_zip, file = "figures/fig_s_ldizip_v2.png", bg = "white")
```
part B. tick suitability
```{r}

tick_inf <- rast("data/raw/ticks/ipac_year_infection_predictions.tif") %>% 
  project("EPSG:4269") 

ldi <- st_read("data/raw/LD_incidence/CA_Zips_Lyme_NAD83.shp") 

ca_zipcode_shp <- ldi %>% 
  st_transform(st_crs(ca_shp)) %>% 
  dplyr::select(ZIP_CODE,PO_NAME, geometry, LDI_Cat)
  #dplyr::select(ZIP_CODE, PO_NAME, LDI_Cat,  GEOID, NAME)


zipcode_tick_df <- read.csv(file = "data/processed/zipcode_tickinf.csv") %>% 
  mutate(ZIP_CODE = as.character(ZIP_CODE),
         ZIP_CODE = str_pad(ZIP_CODE, width = 5, side = "left", pad = "0")) %>% 
  dplyr::select(-X) 
  

zipcode_tick_shp <- ca_zipcode_shp %>% 
  left_join(zipcode_tick_df, by = "ZIP_CODE")
  
map_inftick_zip <-ggplot() +
  geom_sf(data = zipcode_tick_shp, aes(fill = tick_year_inf_prob), color = NA, size = 0.1) +
  scale_fill_viridis_c(option = "H", direction = 1, na.value = "grey90", name = "Infected \nTick\nSuitability") + 
  #labs(tag = "B") +
  theme_void() +
    theme(plot.title = element_text(face = "bold", size = 14), #hjust = 0.5,
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "right",#"bottom",# 
        legend.direction = "vertical", #"horizontal",# 
        #legend.key.width = unit(.25, "cm"), 
        #legend.key.height = unit(.25, "cm"),
        legend.title = element_text(vjust = .7,face = "bold")) 
```

```{r}
zip_activity_annual_agg <- read.csv("data/processed/meta_annual_zip.csv")

# Load and process ZIP code data
ca_zipcode_shp <- zip_activity_annual_agg %>% 
  mutate(
    ZIP_CODE = as.character(ZIP_CODE),
    ZIP_CODE = str_pad(ZIP_CODE, width = 5, side = "left", pad = "0")
  ) %>% 
  dplyr::select(-X) %>% 
  left_join(
    zipcode_tick_shp %>% dplyr::select(ZIP_CODE, geometry), 
    by = "ZIP_CODE"
  ) %>% 
  st_as_sf() %>%                     # ensure it's an sf object
  st_transform(st_crs(ca_shp))



map_humanactiv_zip <- ggplot() +
  geom_sf(data = ca_zipcode_shp, aes(fill = log(total_activity+1)), color = NA, size = 0.1) +
  scale_fill_viridis_c(option = "H", direction = 1, na.value = "grey90", name = "Relative\nHuman\nActivity") + 
  #labs(tag = "C") +
  theme_void() +
    theme(plot.title = element_text(face = "bold", size = 14), #hjust = 0.5,
        plot.margin = margin(t = 1, r = 0, b = 0, l = 0),
        legend.position = "right",#"bottom",# 
        legend.direction = "vertical", #"horizontal",# 
        #legend.key.width = unit(.25, "cm"), 
        #legend.key.height = unit(.25, "cm"),
        legend.title = element_text(vjust = .7,face = "bold")) 

```

combine them
```{r}
fig_s4 <- map_ldi_zip + map_inftick_zip + map_humanactiv_zip + 
  plot_layout(ncol = 3, nrow = 1, heights = c(1.2, 1.2, 1.2)) 

ggsave(fig_s4, file = "figures/updated/fig_s4_v2.png", width = 16, height = 6, bg = "white", dpi = 300)

```


##Supplemental Figure 5. Maps of tick suitability predictions
```{r}
## write function to call in all tiff files
tick_month_tiff <- list.files("data/raw/ticks/ipac_month_suitability_predictions/", full.names = TRUE)
month_names <- month.abb[1:length(tick_month_tiff)]

## fix park crs
sample_rast <- rast(tick_month_tiff[1])

casp_shp_vect_suit_monthly <- casp_shp_clean %>% 
  st_make_valid() %>% 
  st_simplify(dTolerance = 0.001, preserveTopology = TRUE) %>% 
  st_transform(crs = crs(sample_rast)) %>% 
  vect()


## define plotting function
plot_tick_suitability <- function(tif_path, month_label, shp_vect) {
  rast_month <- rast(tif_path)
  
  # aggregate for mmory 
  
  rast_month <- aggregate(rast_month, fact = 5, fun = mean)
  
  # convert to df for ggplot
  df <- as.data.frame(rast_month, xy = TRUE, na.rm = FALSE)
  names(df)[3] <- "tick_inf"
  
  # convert to sf for plotting 
  shp_sf <- st_as_sf(shp_vect)
  
  ggplot() +
    geom_raster(data = df, aes(x = x, y = y, fill = tick_inf)) +
    geom_sf(data = shp_sf, fill = "tomato", color = "tomato", size = 0.6) +
    #geom_sf(data = shp_sf, fill = NA, color = "black", size = 0.6) +
    scale_fill_viridis_c(na.value = "grey95", option = "mako", direction = -1) +
    coord_sf() +
    theme_void() +
    labs(title =  month_label, fill = "Probability") +
    theme(plot.title = element_text(face = "bold", size = 20),
      plot.margin = margin(t = 0, r = 0, b = 0, l = 0),
      legend.position = "bottom",
      legend.direction = "horizontal",
      legend.key.width = unit(1, "cm"),
      legend.title = element_text(vjust = .7)) +
    guides(fill = FALSE)
}


# generate a list of plots
tick_monthly_plots <- map2(.x = tick_month_tiff,
                           .y = month_names,
                           .f = ~plot_tick_suitability(.x, .y, shp_vect = casp_shp_vect_suit_monthly))


# make grid layout
wrap_plots(tick_monthly_plots, ncol = 4) +
  plot_annotation(title = "Monthly Environmental Suitability for Ticks")

dir.create("figures/monthly_maps_v3", showWarnings = FALSE)
for (i in seq_along(tick_month_tiff)) {
  plot_i <- plot_tick_suitability(tif_path = tick_month_tiff[i],
                                  month_label = month_names[i],
                                  shp_vect = casp_shp_vect_suit_monthly)
  
  ggsave(filename = paste0("figures/monthly_maps_v3/", month_names[i], ".png"),
    plot = plot_i, width = 6,height = 5, dpi = 150)
  
  rm(plot_i)
  gc()
}

```

arrange
```{r}
library(magick)
library(cowplot)
library(grid)

# Get PNG paths
img_paths <- list.files(
  "figures/monthly_maps_v3",
  pattern = "\\.png$",
  full.names = TRUE
)

# Optional: sort by month
img_paths <- img_paths[
  match(month.abb, tools::file_path_sans_ext(basename(img_paths)))
]

# Read + aggressively trim
plots <- lapply(img_paths, function(p) {
  img <- image_read(p) |>
    image_trim(fuzz = 40)   # trims transparent + near-white margins

  ggdraw() +
    draw_image(img,  scale = 1)
})

monthly_grid <- plot_grid(
  plotlist = plots,
  ncol = 4,
  align = "hv"
)

final_plot <- plot_grid(
  ggdraw() +
    draw_label(
      "Tick Suitability by Month",
      fontface = "bold",
      size = 18,
      x = 0.5,
      hjust = 0.5
    ),
  monthly_grid,
  ncol = 1,
  rel_heights = c(0.08, 1)
)

ggsave(
  "monthly_maps_grid_v4.png",
  final_plot,
  width  = 12,   # flexible
  height = 9,    # flexible
  dpi = 300,
  bg = "white"
)



```