---
title: "Meta pipeline (new)"
output: html_document
date: "2025-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

meta activity values are normalized per home pixel so to get actual visitation, you need to multiple each visit fraction by the population of home pixel



Step 1. Pull down, filter, and wrangle individual csvs
```{r}
## set up params
chunk_size <- 500000 # number of lines to read in per chunk (helps with memory)
input_file <- "data/raw/recreation/meta/activitymaps/activity_space_distributions_20250908_us1.csv"
output_file <- "data/processed/meta_activitymaps_us1_20251007.csv"

# filter for california (currently this is all of US)
# looked at casp_shp_bbox and added + 0.2
lat_min <- 32.3 
lat_max <- 42.1
lon_min <- -125 # western boundary
lon_max <- -114 # eastern boundary

# choose zoom level (larger zoom values = smaller tiles)
zoom <- 13 # may fiddle with this (10 is way too coarse for me)

## deleting exisiting output file (helps memory)
# initial output - makes we we dont append to old runs results
if(file.exists(output_file)) file.remove(output_file)

## read and filter csv into chunks
con <- file(input_file, "r") # open connection to large csv
header <- readLines(con, n = 1) # read first line w/header

## process the csv in chunks (helps memory)
while(length(chunk_lines <- readLines(con, n = chunk_size)) > 0 ) {
  
  # combine header with current chunk lines for data table
  chunk <- fread(paste(c(header, chunk_lines), collapse = "\n"))
  
  # filter chunk for rows that fall within CA boundary
  chunk[
    visit_latitude >= lat_min & visit_latitude <= lat_max &
      visit_longitude >= lon_min & visit_longitude <= lon_max
  ][
    , fwrite(.SD, output_file, append = TRUE) # write only filtered data
  ]
}

close(con) # close file connection

# load filtered data into memory (vs load in full lg csv)
ca_data_1 <- fread(output_file) # read.csv("data/processed/meta_activitymaps_us1_20251007.csv")
#ca_data_2 <- fread(output_file) # "data/processed/meta_activitymaps_us2_20251007.csv"

ca_data_1 <- fread(output_file) #read.csv("data/processed/meta_activitymaps_us1_20251007.csv")
#ca_data_2 <- fread("data/processed/meta_activitymaps_us2_20251007.csv")


```

join home population density
```{r}
gpw <- rast("data/raw/gpw-v4-population-density-rev11_2020_30_sec_asc/gpw_v4_population_density_rev11_2020_30_sec_1.asc")

## for ca 1
home_coords1 <- ca_data_1 %>%
  filter(day_or_night == "daytime") %>%
  distinct(home_latitude, home_longitude) %>%
  mutate(home_lat_round = round(home_latitude, 5),
         home_lon_round = round(home_longitude, 5))


home_sf1 <- home_coords1 %>% 
  st_as_sf(coords = c("home_lon_round", "home_lat_round"), crs = 4326) 


home_sf_proj1 <- st_transform(home_sf1, crs(gpw))


# Extract population
home_sf_proj1$home_population <- terra::extract(gpw, vect(home_sf_proj1))[,2]


### for ca 2
home_coords2 <- ca_data_2 %>%
  filter(day_or_night == "daytime") %>%
  distinct(home_latitude, home_longitude) %>%
  mutate(home_lat_round = round(home_latitude, 5),
         home_lon_round = round(home_longitude, 5))


home_sf2 <- home_coords2 %>% 
  st_as_sf(coords = c("home_lon_round", "home_lat_round"), crs = 4326) 


home_sf_proj2 <- st_transform(home_sf2, crs(gpw))


# Extract population
home_sf_proj2$home_population <- terra::extract(gpw, vect(home_sf_proj2))[,2]

```

now add home pop and summarize


```{r}
ca_data_1 <- ca_data_1 %>% 
  left_join(home_sf_proj1, by = c("home_latitude", "home_longitude")) %>% 
  mutate(home_population = ifelse(is.na(home_population), 0, home_population),
         weighted_activity = home_population*visit_fraction)


ca_data_2 <- ca_data_2 %>% 
  left_join(home_sf_proj2, by = c("home_latitude", "home_longitude")) %>% 
  mutate(home_population = ifelse(is.na(home_population), 0, home_population),
         weighted_activity = home_population*visit_fraction)


## summmarize each visit pixel
activity_summary_1 <- ca_data_1 %>% 
  group_by(visit_latitude, visit_longitude) %>% 
  summarise(total_weighted_activity = sum(weighted_activity, na.rm = TRUE), .groups = "drop")


activity_summary_2 <- ca_data_2 %>% 
  group_by(visit_latitude, visit_longitude) %>% 
  summarise(total_weighted_activity = sum(weighted_activity, na.rm = TRUE), .groups = "drop")

## combine and collapse to unique visit pixels
activity_summary_12  <- bind_rows(activity_summary_1, activity_summary_2) %>% 
  group_by(visit_latitude, visit_longitude) %>% 
  summarise(total_weighted_activity = sum(total_weighted_activity, na.rm = TRUE), .groups = "drop") # 9224

```


visualize

```{r}
activity_sf <- st_as_sf(
  activity_summary_2,
  coords = c("visit_longitude", "visit_latitude"),
  crs = 4326)

pal <- colorNumeric(
  palette = "YlOrRd",  # You can change this to "Viridis", "Blues", etc.
  domain = log(activity_sf$total_weighted_activity+1),
  na.color = "transparent",
  reverse = FALSE) 

leaflet(activity_sf) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = 5,
    fillColor = ~pal(log(total_weighted_activity+1)),
    fillOpacity = 0.8,
    color = "black",
    weight = 0.2,
    popup = ~paste0("Total Weighted Activity: ", round(total_weighted_activity, 2))
  ) %>%
  addLegend(
    pal = pal,
    values = ~log(total_weighted_activity+1),
    title = "log(Pop-Weighted Activity+1)",
    position = "bottomright"
  )

activity_summary_2

activity_summary12 #16,934 × 3
```

convert back to tiles so there's more coverage
```{r}
## compute activity indices for each tile location
tile_df <- activity_summary_12 %>% 
  rowwise() %>% # need for lonlat_* since its returned as a list per row
  mutate(tile = list(lonlat_to_tilenum(visit_longitude, visit_latitude, zoom))) %>% 
  mutate(xtile = tile$x, ytile = tile$y) %>% # extract x & y index of tile
  ungroup()

## aggregate activity by tile 
# remember i'm using max here but could be mean
tile_activity <- tile_df %>% 
  group_by(xtile, ytile) %>% 
  summarise(activity = sum(total_weighted_activity), .groups = "drop")

## compute bounding box for each tile
# for each tile (xtile, ytile) comput its bbox in lat/lon coords
bbox_df <- tile_activity %>% 
  rowwise() %>% 
  mutate(bbox = list(tile_bbox(xtile, ytile, zoom))) %>% 
  mutate(lng1 = bbox$xmin, lng2 = bbox$xmax, # left/right longitudes
         lat1 = bbox$ymin, lat2 = bbox$ymax) %>%  # bottom/top latitudes
  ungroup()

## create tile polygons
# build each tile polygon using its 4 corners, and close it by repeating the first point
polygon_list <- purrr::pmap(list(bbox_df$lng1, bbox_df$lng2, bbox_df$lat1, bbox_df$lat2),
                            function(xmin, xmax, ymin, ymax) {
                              st_polygon(list(matrix(c(
                                 xmin, ymin, # bottom left
                                 xmax, ymin, # bottom eirght
                                 xmax, ymax, # top right
                                 xmin, ymax, # top left
                                 xmin, ymin), # close polygon
                                 ncol = 2, byrow = TRUE)))
                            })

## create sf object
# attach activity values & geometry into a spatial object
tile_activity_sf <- st_sf(activity = bbox_df$activity,
                          geometry = st_sfc(polygon_list, crs = 3857)) # check, but i thik tile_bbox gives web mercator coords

# convert 3857 to 4326 for leaflet & downstream analysis
tile_activity_lonlat <- st_transform(tile_activity_sf, crs = 4326)

#activitymap_us1 <- tile_activity_lonlat # will switch this for the next csv
 #activitymap_us2 <- tile_activity_lonlat
activitymap_us12 <- tile_activity_lonlat

activitymap_us12 <- activitymap_us12 %>% mutate(log_activity = log(activity+1))
#st_write(activitymap_us12, "data/processed/meta_activitymaps_us12.shp")
#st_write(activitymap_us1, "data/processed/meta_activitymaps_us1.shp") # i put these into folders _shp
#st_write(activitymap_us2, "data/processed/meta_activitymaps_us2.shp")

# create color scale
pal <- colorNumeric("YlOrRd", domain = log(tile_activity_lonlat$activity+1), na.color = "transparent")

## create leaflet map
leaflet(tile_activity_lonlat) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~pal(log(activity+1)),
              weight = 1, color = "black", fillOpacity = 0.7,
              popup = ~paste0("log(Activity+1)", round(log(activity+1), 3))) %>% 
  addLegend(pal = pal, values = ~log(activity+1), title = "log(Activity Level+1)", position = "bottomright")


```


NOW RERUN ALL ANALYSES AND FIGURES WITH THIS!!! EXTRACT LOG ACTIVITY TO PARK POLYGON

